Categories of Optics

### 2018-07-30 paper.tex
State of play

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Twoptic}{\mathbf{Optic}^2}
\newcommand{\Lawful}{\mathbf{Lawful}}
%\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Setter}{\mathbf{Setter}}
\newcommand{\Traversal}{\mathbf{Traversal}}
\newcommand{\Getter}{\mathbf{Getter}}
\newcommand{\Review}{\mathbf{Review}}
\newcommand{\Fold}{\mathbf{Fold}}

@@

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}
\newcommand{\inl}{\mathrm{inl}}
\newcommand{\inr}{\mathrm{inr}}
\DeclareMathOperator{\im}{im}
\newcommand{\act}{\cdot}
\newcommand{\codisc}{\mathsf{codisc}}

@@

In its most concrete form, a \emph{lens} from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$. From a software engineering standpoint, such a lens allows us to ``zoom in'' on $S$ to focus on a small part $A$, manipulate $A$ in some way, then ``zoom out'' and have our changes reflected in $S$.

Often we want our lens to obey the following laws:
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be represented faithfully in $S$;
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}
We will call such lenses \emph{lawful}. (The names of the laws make more sense when the equations are written with composition in diagrammatic order!) Lenses form a category with a subcategory of lawful lenses.

%\todo{reference a whole bunch of bx and database stuff}

Lenses were discovered to be just one of a hierarchy of data accessors, including prisms, setters, traversals, and more, collectively called \emph{optics}. Each optic variant has a concrete description as a certain collection of maps, with attendant laws under which we consider them well-behaved. We begin in Section~\ref{sec:optics} by defining the \emph{category of optics} in a sufficiently general way to encompass almost all the optic variants in use in the wild, using lenses as a running example. Section~\ref{sec:lawful-optics} defines the equivalent of the lens laws for a general category of optics. Then in Section~\ref{sec:examples} we see that these generic definitions specialise correctly to the other basic varieties of optic, including giving the correct laws.

@@

@@

We begin by defining the category of optics in a symmetric monoidal category $\C$. This category was first defined in~\cite[Section 6]{Doubles} as the `double' of a monoidal category. There it was used for a completely different purpose---to investigate the relationship between Tambara modules and the `center' of a monoidal category. Our definition is identical, other than flipping the direction of the morphisms to match the existing work on lenses, and restricting our attention to the unenriched setting.

Our definition of optic has as domain and codomain \emph{pairs} of objects of $\C$. This generality is important for the applications to game theory; for ordinary lenses we restrict our attention to pairs with the same object in both entries.

@@

Because this coend takes place in $\Set$, we can use Lemma~\ref{lemma:calculate-coend} to describe $\Optic_\otimes((S, S'), (A, A'))$ explicitly. It is the set of pairs $(l, r)$, where $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, quotiented by the relation
\begin{align*}
  ((f \otimes A) l, r) \sim (l, r (f \otimes A'))
\end{align*}
for any $l : S \to M \otimes A$, $r : N \otimes A' \to S'$ and $f : M \to N$.

For a pair of maps $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\rep{l}{r} : (S, S') \hto (A, A')$ for their image in $\Optic_\otimes((S, S'), (A, A'))$, and say that $M$ is the residual for this representative. Optics will always be written with a crossed arrow $\hto$ to distinguish them from morphisms of $\C$.

@@

\begin{remark}
  Elements of $\Optic_\otimes((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\rep{l}{r}$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation imposed by the coend equates the following two diagrams:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \qquad
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut point between $l$ and $r$ in most subsequent diagrams; any choice yields a representative of same optic.
\end{remark}

@@

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
                                                                                                                                                                       \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact and the Fubini theorem for coends, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
natural in $M$ and $N$. For these we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\rep{l'}{r'} : (R, R') \hto (S, S')$ and $\rep{l}{r} : (S, S') \hto (A, A')$ with $M$ the residual for $\rep{l'}{r'}$. The composite is then: $\rep{l}{r} \circ \rep{l'}{r'} = \rep{(M \otimes l)l'}{r'(M \otimes r)}$.

Graphically, this is substituting the first optic for the hole of the second:
\begin{center}
  \input{diagrams/generic-optic-noline.tikz}
  \quad\raisebox{1.5cm}{$\circ$}\qquad
  \input{diagrams/generic-optic2-noline.tikz} \\
  \raisebox{1.5cm}{$:=$}\qquad
  \input{diagrams/optic-composition.tikz}
\end{center}

@@

Using the isomorphism in both directions, we leave it to the reader to verify that composition in $\Optic_\times$ corresponds to ordinary composition of concrete lenses. (Of course, there is only one possible way such a collection of morphisms could be sensibly composed!)
\end{remark}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $M \act A$ or $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.
\end{definition}

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $- \act A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
  \begin{align*}
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}} \\
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
  \end{align*}

\begin{remark}
It is shown in~\cite[Section 6]{ANoteOnActions} that, at least in the case $\M$ is right closed, to give such an action is equivalent to giving a ``copowered $\M$-category $\C$''. \todo{Pointed out to me years ago when I emailed Ross Street, but I didn't understand what he was saying}
\end{remark}

@@

\begin{itemize}
\item \emph{``Achromatic'' Lenses}~\cite[Section 5.2]{ProfunctorOpticsThesis} are lenses that also admit an operation $\fcreate : A \to S$. These are optics for the action of $\C$ on itself by $M \act A = (M \sqcup 1) \times A$, or equivalently, of the category of pointed objects of $\C$ on $\C$ by cartesian product. Concrete achromatic lenses $(S, S') \hto (A, A')$ are elements of the set \[\C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S').\]
%\begin{align*}
%  \Optic((S, S'), (A, A'))
%  &= \int^{M \in \C} \C(S, M \act A) \times \C(M \act A', S') \\
%  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
%  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
%  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
%\end{align*}
\item \emph{Affine Traversals}~\cite{AffineTraversalPost} allow access to a target that may or may not be present. Suppose $\C$ is cartesian closed and has binary coproducts. Let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
The category $\mathsf{Aff}$ acts on $\C$ by $(P, Q) \act A = P \sqcup (Q \times A)$, in fact, $\mathsf{Aff}$ is cooked up exactly as the closure of the actions $- \times A$ and $- \sqcup A$ under composition. A concrete affine traversal is an element of \[\C(S, S' \sqcup (\homC(A', S') \times A)).\]

@@

We can generalise the definition of $\Optic$ so that the two halves lie in different categories, allowing us to capture a few more optic variants found in the wild. Suppose $\C_1$ and $\C_2$ are categories that are each acted on by a common monoidal category $\M$.

\begin{definition}
  Given two objects of $\C_1 \times \C_2^\op$, say $(S, S')$ and $(A, A')$, an \emph{optic} $p : (S, S') \hto (A, A')$ is an element of the set
  \begin{align*}
    \Optic_\otimes((S, S'), (A, A')) := \int^{M \in \M} \C_1(S, M \act A) \times \C_2(M \act A', S')
  \end{align*}
\end{definition}

Several results from Section~\ref{sec:optics} continue to hold:
\begin{proposition}\label{prop:mixed-optic-is-cat}
  The above data form a category $\Optic_{\M, \C_1, \C_2}$. \todo{notation?} \qed
\end{proposition}
\begin{proposition}
  There is a functor $\iota : \C_1 \times \C_2^\op \to \Optic_{\M, \C_1, \C_2}$, which on objects is given by $\iota(S, S') = (S, S')$ and on a morphism $(f, g) : (S, S') \to (A, A')$ by \[\iota(f, g) = \rep{(\lambda^{\C_1}_A)^{-1} f}{g \lambda^{\C_2}_{A'}},\] where $\lambda^{\C_1}_A : A \to IA$ in $\C_1$ and $\lambda^{\C_2}_{A'} : A' \to IA'$ in $\C_2$. \qed
\end{proposition}
\begin{proposition}\label{prop:mixed-change-of-action}
  Given two morphisms of actions $F_1 : (\M, \C_1) \to (\N, \D_1)$ and $F_2 : (\M, \C_2) \to (\N, \D_2)$, there is an induced functor $\Optic(F_1, F_2) : \Optic_{\M, \C_1, \C_2} \to \Optic_{\N, \D_1, \D_2}$. \qed
\end{proposition}

The notion of lawfulness used earlier breaks down in this setting, as the definition of $\Twoptic(S, A)$ only makes sense when $\C_1 = \C_2$. The following definition gives a reasonable approximation.

\begin{definition}
If $F_1 : (\M, \C_1) \to (\N, \D)$ and $F_2 : (\M, \C_2) \to (\N, \D)$, a mixed optic $p : (S, S) \hto (A, A)$ in $\Optic_{\M, \C_1, \C_2}$ is \emph{lawful relative to $(F_1, F_2)$} if its image under $\Optic(F_1, F_2)$ is lawful.
\end{definition}

\subsection{Examples}

\subsubsection{Degenerate Optics}

@@

Now for a category with finite products $\C$, define $\Getter$ to be the category of mixed optics $\Optic_{\times, \C, \C_\codisc}$. This is indeed isomorphic to $\C$:
\begin{align*}
&\int^{M \in \M} \C(S, M \times A) \times \C_\codisc(M \times A', S')\\
&\cong \int^{M \in \M} \C(S, M \times A) \\
&\cong \C(S, 1 \times A) \\
&\cong \C(S, A)
\end{align*}
This works for $\Review := \Optic_{\sqcup, \C_\codisc, \C}$ in the same way. There are induced functors $\Optic(\id_\C, i) : \Lens \to \Getter$ and $\Optic(i, \id_\C) : \Prism \to \Review$.

@@

\subsubsection{Indexed Optics}
As well as giving a principled home to degenerate optics, the notion of mixed optics also allows us to describe \emph{indexed optics} categorically. The most interesting example of an indexed optic variant is the \emph{indexed traversal}, which, for example, allows access to the indices of the entries of a list as it is being traversed.

We first need to define indexed traversable functors. Unfortunately this has many moving parts!

\begin{proposition}
The assignment $F \act X \mapsto F \Set(I, X)$ is a lax monoidal action of $\App$ on $\Set$, called the \emph{$I$-indexed action of $\App$ on $\Set$}.
\end{proposition}

\begin{definition}
A \emph{traversable functor indexed by $I$} is a functor $T : \Set \to \Set$ equipped with a distributive law $\delta_F : T(\Set(I, F-)) \to FT$ for $T$ between the $I$-indexed action of $\App$ on $\Set$ and the action of $\App$ on $\Set$ by evaluation.
\end{definition}

\begin{definition}
The category of \emph{indexed traversable functors} $\IdxTraversable$ has as objects pairs $(T, I)$, where $I \in \Set$ and $T : \Set \to \Set$ is a traversable functor indexed by $I$. A morphism $(T, I) \to (T', I')$ is a natural transformation $\alpha : T \to T'$ and morphism (note the direction!) $i : I' \to I$, such that $\alpha$ commutes with the distributive laws on both traversable functors, using $i$ to mediate between the index objects.
\end{definition}

\begin{proposition}
$\IdxTraversable$ acts on $\C \times \C$ by $(T, I) \act (A, R) \mapsto (TA, \homC(I, R))$ and on $\C$ simply by $(T, I) \act A \mapsto TA$.
\end{proposition}

>>>>

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Twoptic}{\mathbf{Optic}^2}
\newcommand{\Lawful}{\mathbf{Lawful}}
%\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Setter}{\mathbf{Setter}}
\newcommand{\Traversal}{\mathbf{Traversal}}
\newcommand{\Getter}{\mathbf{Getter}}
\newcommand{\Review}{\mathbf{Review}}
\newcommand{\Fold}{\mathbf{Fold}}
\newcommand{\IdxLens}{\mathbf{IdxLens}}
\newcommand{\IdxTraversal}{\mathbf{IdxTraversal}}

@@

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}
\newcommand{\inl}{\mathrm{inl}}
\newcommand{\inr}{\mathrm{inr}}
\DeclareMathOperator{\im}{im}
\newcommand{\act}{\cdot}
\newcommand{\codisc}{\mathsf{codisc}}
\DeclareMathOperator*{\colim}{\mathrm{colim}}

\newcommand*\circled[1]{\tikz[baseline={([yshift=-0.65ex]current bounding box.center)}]{
   \node[shape=circle,draw,inner sep=1pt] (char) {#1};}}
\newcommand{\actL}{{\circled{\tiny$\mathsf{L}$}}}
\newcommand{\actR}{{\circled{\tiny$\mathsf{R}$}}}

@@

In its most concrete form, a \emph{lens}~\cite{CombinatorsForBidirectionalTreeTransformations} from a set $S$ to a set $A$ is a pair of maps ${\fget : S \to A}$ and ${\fput : S \times A \to S}$. From an engineering standpoint, such a lens allows us to ``zoom in'' on $S$ to focus on a small part $A$, manipulate $A$ in some way, then ``zoom out'' and have our changes reflected in $S$.

So that our lenses better conform to this intuitive idea of ``zooming in'', we often want them satisfy some conditions known as the \emph{lens laws}:
\begin{center}
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[ 
\begin{tikzcd}
S \times A \ar[rr, "\fput"] \ar[dr, "\pi_2", swap] && S \ar[dl, "\fget"] \\
& A
\end{tikzcd}
\]
\hspace{0.8cm}$\fput\fget$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[ 
\begin{tikzcd}
S \ar[rr, "{[\id_S, \fget]}"] \ar[dr, "\id_S", swap] && S \times A \ar[dl, "\fput"] \\
& A
\end{tikzcd}
\]
\hspace{-0.8cm}$\fget\fput$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[ 
\begin{tikzcd}
S \times A \times A \ar[r, "\fput \times A"] \ar[d, "\pi_{1, 3}", swap] & S \times A\ar[d, "\fput"] \\
S \times A \ar[r, "\fput", swap] & A
\end{tikzcd}
\]
\quad$\fput\fput$
\end{center}
\end{minipage}%
\end{center}
The $\fput\fget$ law states that any update to $A$ is represented faithfully in $S$. The $\fget\fput$ law states that if $A$ is not changed then neither is $S$; and finally, the $\fput\fput$ law states that any update to $A$ completely overwrites previous updates.

%Lenses form a category, with the composition of lenses $T \hto S$ and $S \hto A$ as indicated:
%\begin{align*}
%\fget &: T \xrightarrow{\fget_1} S \xrightarrow{\fget_2} A \\
%\fput &: T \times A \xrightarrow{[\id_T, \fget_1, \id_A]} T \times S \times A \xrightarrow{T \times \fput_2} T \times S \xrightarrow{\fput_1} T
%\end{align*}
%This composition preserves lawfulness.
%\todo{reference a whole bunch of bx and database stuff}

Lenses were discovered to be just one of a hierarchy of data accessors, including prisms, setters, traversals, and more, collectively called \emph{optics}. Each optic variant has a concrete description as a certain collection of maps, with attendant laws under which we consider them well-behaved. We begin in Section~\ref{sec:optics} by defining the \emph{category of optics} in a sufficiently general way to encompass almost all the optic variants in use in the wild, using lenses as a running example. Section~\ref{sec:lawful-optics} defines the equivalent of the lens laws for a general category of optics. Then in Section~\ref{sec:examples} we see that these generic definitions specialise correctly to the other basic varieties of optic, including giving the correct laws. We include Section~\ref{sec:mixed-optics} a more speculative discussion of ``mixed optics'', which include indexed and coindexed lenses and traversals.

@@

\begin{lemma}[Mute coends]
Consider a functor $F : \C \to \E$ as a functor $\C^\op \times \C \to \E$ that ignores its contravariant argument. Then \[ \int^C F(C) \cong  \colim F. \] \qed
\end{lemma}

@@

We begin by defining the category of optics for a symmetric monoidal category $\C$. This category was first defined in~\cite[Section 6]{Doubles} as the `double' of a monoidal category. There it was used for a completely different purpose---to investigate the relationship between Tambara modules and the `center' of a monoidal category. Our definition is identical, other than flipping the direction of the morphisms to match the existing work on lenses, and restricting our attention to the unenriched setting.

Our definition of optic has as domain and codomain \emph{pairs} of objects of $\C$, one of which behaves covariantly and the other contravariantly. For example, our lenses will be pairs of maps $\fget : S \to A$ and $\fput : S \times A' \to S'$. This generality is important for the applications to game theory, and in fact helps in calculations by making the covariant and contravariant positions more difficult to confuse. Readers more familiar with lenses should ignore the primes.

@@

Because this coend takes place in $\Set$, we can use Lemma~\ref{lemma:calculate-coend} to describe $\Optic_\otimes((S, S'), (A, A'))$ explicitly. It is the set of pairs $(l, r)$, where $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, quotiented by the equivalence relation generated by relations of the form
\begin{align*}
  ((f \otimes A) l, r) \sim (l, r (f \otimes A'))
\end{align*}
for any $l : S \to M \otimes A$, $r : N \otimes A' \to S'$ and $f : M \to N$.

For a pair of maps $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\rep{l}{r} : (S, S') \hto (A, A')$ for their image in $\Optic_\otimes((S, S'), (A, A'))$, and say that the object $M$ is the residual for this representative. Optics will always be written with a crossed arrow $\hto$ to distinguish them from morphisms of $\C$.

@@

\begin{remark}
  Elements of $\Optic_\otimes((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\rep{l}{r}$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation imposed by the coend can be graphically expressed as:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \qquad
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut between $l$ and $r$ in most subsequent diagrams; any choice yields a representative of same optic.
\end{remark}

@@

Two optics can be composed as follows. The easiest interpretation is graphical: composition of optics corresponds to substituting the first optic for the hole of the second:
\begin{center}
  \input{diagrams/generic-optic-noline.tikz}
  \quad\raisebox{1.5cm}{$\circ$}\qquad
  \input{diagrams/generic-optic2-noline.tikz} \\
  \raisebox{1.5cm}{$:=$}\qquad
  \input{diagrams/optic-composition.tikz}
\end{center}

More formally, we wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{N \in \C} \C(R, N \otimes S) \times \C(N \otimes S', R')\right) \\ 
  &\quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact and the Fubini theorem for coends, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
natural in $M$ and $N$. For these we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
Written equationally, suppose $\rep{l'}{r'} : (R, R') \hto (S, S')$ and $\rep{l}{r} : (S, S') \hto (A, A')$ are optics with $M$ the residual for $\rep{l'}{r'}$. The composite $(R, R') \hto (A, A')$ is then: \[\rep{l}{r} \circ \rep{l'}{r'} := \rep{(M \otimes l)l'}{r'(M \otimes r)}.\]

@@

Using the isomorphism in both directions, we leave it to the reader to verify that composition in $\Optic_\times$ corresponds to ordinary composition of concrete lenses. (Of course, there is only one sensible way to compose such a collection of morphisms!)
\end{remark}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $M \act A$ or $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.

An action is \emph{lax} if $a$ is instead a lax monoidal functor.
\end{definition}

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $- \act A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
    \begin{equation*}
    \begin{aligned}[t]
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}}
    \end{aligned}
    \qquad\qquad\qquad
    \begin{aligned}[t]
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
    \end{aligned}
\end{equation*}
\begin{remark}
It is shown in~\cite[Section 6]{ANoteOnActions} that, at least in the case $\M$ is right closed, to give such an action is equivalent to giving a ``copowered $\M$-category $\C$''. \todo{Pointed out to me years ago when I emailed Ross Street, but I didn't understand what he was saying}
\end{remark}

@@

\begin{itemize}
\item \emph{``Achromatic'' Lenses}~\cite[Section 5.2]{ProfunctorOpticsThesis} are lenses that also admit an operation $\fcreate : A \to S$. These are optics for the action of $\C$ on itself by $M \act A = (M \sqcup 1) \times A$, or equivalently, of the category of pointed objects of $\C$ on $\C$ by cartesian product. Concrete achromatic lenses $(S, S') \hto (A, A')$ are elements of the set \[\C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S').\]
%\begin{align*}
%  \Optic((S, S'), (A, A'))
%  &= \int^{M \in \C} \C(S, M \act A) \times \C(M \act A', S') \\
%  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
%  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
%  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
%\end{align*}
\item \emph{Affine Traversals}~\cite{AffineTraversalPost} allow access to a target that may or may not be present. Suppose $\C$ is cartesian closed and has binary coproducts. Let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
The category $\mathsf{Aff}$ acts on $\C$ by $(P, Q) \act A = P \sqcup (Q \times A)$, in fact, $\mathsf{Aff}$ is cooked up to act on $\C$ exactly by the closure of the actions $- \times A$ and $- \sqcup A$ under composition. A concrete affine traversal is an element of \[\C(S, S' \sqcup (\homC(A', S') \times A)).\]

@@

\todo{This bit is very much in progress, skip to Section~\ref{sec:profunctor-optics}}

We can generalise the definition of $\Optic$ so that the two halves lie in different categories, allowing us to capture a few more optic variants found in the wild. Suppose $\C_L$ and $\C_L$ are categories that are each acted on by a common monoidal category $\M$. Write these actions as $\actL$ and $\actR$ respectively.

\begin{definition}
  Given two objects of $\C_L \times \C_R^\op$, say $(S, S')$ and $(A, A')$, an \emph{optic} $p : (S, S') \hto (A, A')$ is an element of the set
  \begin{align*}
    \Optic_{\actL, \actR}((S, S'), (A, A')) := \int^{M \in \M} \C_L(S, M \actL A) \times \C_R(M \actR A', S')
  \end{align*}
\end{definition}

Several results from Section~\ref{sec:optics} continue to hold:
\begin{proposition}\label{prop:mixed-optic-is-cat}
  $\Optic_{\actL, \actR}$ is a category. \qed
\end{proposition}
\begin{proposition}
  There is a functor $\iota : \C_L \times \C_R^\op \to \Optic_{\actL, \actR}$, which on objects is given by $\iota(S, S') = (S, S')$ and on a morphism $(f, g) : (S, S') \to (A, A')$ by \[\iota(f, g) = \rep{(\lambda^{\C_L}_A)^{-1} f}{g \lambda^{\C_R}_{A'}},\] where $\lambda^{\C_L}_A : A \to IA$ in $\C_L$ and $\lambda^{\C_R}_{A'} : A' \to IA'$ in $\C_R$. \qed
\end{proposition}
\begin{proposition}\label{prop:mixed-change-of-action}
  Given two morphisms of actions $F_L : (\M, \C_L) \to (\N, \D_L)$ and $F_R : (\M, \C_R) \to (\N, \D_R)$, there is an induced functor $\Optic(F_L, F_R) : \Optic_{\actL_1, \actR_1} \to \Optic_{\actL_2, \actR_2}$. \qed
\end{proposition}

The notion of lawfulness used earlier breaks down in this setting, as the definition of $\Twoptic(S, A)$ only makes sense when $\C_L = \C_R$. The following definition gives a reasonable approximation.

\begin{definition}
If $F_L : (\M, \C_L) \to (\N, \D)$ and $F_R : (\M, \C_R) \to (\N, \D)$, a mixed optic $p : (S, S) \hto (A, A)$ in $\Optic_{\actL, \actR}$ is \emph{lawful relative to $(F_L, F_R)$} if its image under $\Optic(F_L, F_R)$ in $\Optic_\N$ is lawful.
\end{definition}

\subsection{Degenerate Optics}

@@

Now for a category with finite products $\C$, define $\Getter$ to be the category of mixed optics $\Optic_{\times, \C, \C_\codisc}$. This is indeed isomorphic to $\C$:
\begin{align*}
&\int^{M \in \M} \C(S, M \times A) \times \C_\codisc(M \times A', S')\\
&\cong \int^{M \in \M} \C(S, M \times A) \\
&\cong \C(S, 1 \times A) \\
&\cong \C(S, A)
\end{align*}
The only sensible notion of lawfulness for $\Getter$s is lawfulness relative to $(i, \id_{\C_\codisc})$, but all optics in $\Optic_{\C_\codisc}$ are trivially lawful.

One defines $\Review := \Optic_{\sqcup, \C_\codisc, \C}$ in the same way. There are induced functors $\Optic(\id_\C, i) : \Lens \to \Getter$ and $\Optic(i, \id_\C) : \Prism \to \Review$.

@@

\subsection{Indexed Lenses}
As well as giving a principled home to degenerate optics, the notion of mixed optics also allows us to describe \emph{indexed optics} categorically. The most interesting example of an indexed optic variant is the \emph{indexed traversal}, which, for example, allows one to know the index of an entry of a list as one is traversing it. As a warmup, however, let us look at \emph{indexed lenses}, which provide access to an index alongside a target, but do not allow updates to the index.

Suppose $\C$ is cartesian closed, and has an initial object. \todo{check which of these we need} We define two actions of $\C \times \C$, the first on $\C$ and the second on $\C \times \C$:
\begin{align*}
(M, R) \actL (A, I) &:= M \times R \times A \\
(M, R) \actR (A, I) &:= (M \times A, R \times I)
\end{align*}
\todo{I could call this the ``indexed product action'' but the term indexed product is already taken of course}

%\begin{definition}
%The category of \emph{indexed lenses} $\IdxLens$ in $\C$ is the category of mixed optics with $\C \times \C^\op$ acting on the left, and the ordinary product action on the right.
%\end{definition}
